A matrix-free approach for finite-strain hyperelastic problems using geometric multigrid

### 2018-11-01 doc/large_strain.tex
referential/spatial configuration

\subsection{Weak form}
%
% Kinematics
%
The deformation of a body $\mcl B$ from the reference configuration $\mcl B_0$ to the current configuration $\mcl B_t$ at time $t$
is defined via the deformation map $\gz \varphi_t: \mcl B_0 \rightarrow \mcl B_t$, which places material points of $\mcl B$ into the Euclidean space $\mathbb E^3$.
The current location of a material particle $\gz X$ is given by $\gz x = \gz \varphi_t (\gz X)$.
The displacement field reads $\gz u = \gz x - \gz X$.

@@

%
% strain measures
%
In what follows, we parameterize the material behavior using the right Cauchy-Green tensor $\gz C := \gz F^T \cdot \gz F$.
We will also use the Green-Lagrange strain tensor $\gz E:=\frac{1}{2}\left[\gz C - \gz I\right]$ and the left Cauchy-Green tensor $\gz b := \gz F \cdot \gz F^T$.
Clearly $J = \sqrt{\det \gz C}$ and $\partial (\bullet)/\partial \gz E = 2\, \partial (\bullet)/\partial \gz C$.
%
For conservative systems without body forces, the total potential energy functional $\mcl E$ is introduced as
\begin{align}
\mcl E =
\int_{\mcl B_0} \mcl \psi(\gz F) \, \d V \,
- \int_{\partial \mcl B_0^N} \overline{\gz T} \cdot \gz u \, \d S \, ,
\end{align}
where $\overline{\gz T}$ is the prescribed loading at the Neumann part of the boundary $\partial \mcl B_0^N$ in the reference configuration, $\mcl \psi$ denotes the strain-energy per unit reference volume and $\gz u$ should satisfy the prescribed Dirichlet boundary conditions $\gz u = \overline{\gz u}$ on $\partial \mcl B_0^D := \partial \mcl B_0 \setminus \partial \mcl B_0^N \neq \emptyset $.

@@

The Newton-Raphson solution approach is adopted for the nonlinear problem considered here.
Given the current trial solution field $\overline{\gz u^h}$, the correction $\Delta \gz u^h$ field is obtained as the solution to the following system of equations (see \eqref{eq:weak_form} and \eqref{eq:tangent}):
\begin{align}
  A_{ij} \Delta u_j &= - F_i  \label{eq:linear_system} \\
  A_{ij} \equiv a(\gz N_i, \gz N_j) &=
  \int_{\mcl B_0}
  \underbrace{
  \overline{\grad^s} \gz N_i : J \boldsymbol{\mathcal{C}} : \overline{\grad^s} \, \gz N_j
  +
  \overline{\grad}\gz N_i :
  \left[
  \overline{\grad} \gz N_j \cdot
  \overline{\gz \tau}
  \right]
  }_{\displaystyle \tilde{a}(\gz N_i, \gz N_j)}
  \d V
  \label{eq:algebraic_tangent}
  \\
  F_i &=
  \int_{\mcl B_0} \overline{\gz \tau} : \grad^{s} \, \gz N_i \, \d V \,
  -
  \int_{\partial \mcl B_0^N} \overline{\gz T} \cdot \gz N_i \, \d S.
\end{align}
Here $\gz A$ is the discrete tangent {\color{red}operator}, $\gz F$ is the discrete gradient
and $\tilde{a}(\gz N_i, \gz N_j)$ is a representation for the integrand in the bilinear form $a(\gz N_i, \gz N_j)$ for the trial solution field $\overline{\gz u^h}$.
Note that $\overline{\grad^s} \gz N_i$ and $\overline{\grad} \gz N_i$ are gradients of shape functions in the deformed configuration, whereas the integration is done in the reference (undeformed) configuration.

@@

Given that the matrix-free operators are applied on-the-fly, it is crucial to find a good balance between pre-calculation of certain values (caching) and performance of the algorithm.
Below, we propose three algorithms to implement the tangent {\color{red}operator} of finite-strain elasticity (see \eqref{eq:algebraic_tangent}) using the matrix-free approach.
%
In order to obtain gradients with respect to the deformed configuration we used the \textit{MappingQEulerian} class from the \texttt{deal.II} \cite{dealII90} library.
In addition to the standard mapping from the reference (isoparametric) element to the element in real space, this class computes the mapping to the deformed configuration given a displacement field.
%
Due to the specifics of matrix-free operator evaluation implemented in \texttt{deal.II}, the integration is performed over the deformed configuration. Therefore, we have to divide by the Jacobian $J$ of the deformation map at a given quadrature point additionally to express the integral in the undeformed configuration.

@@

Algorithm \ref{alg:mf_scalar} caches a single scalar which involves the logarithm of the Jacobian -- a computationally demanding operation compared to additions and multiplications. As evident from Section \ref{sec:theory}, we need to re-evaluate the Kirchhoff stress at each quadrature point in order to apply the operator. This in turn requires evaluating the deformation gradient $\gz F$ and therefore gradients with respect to the undeformed configuration.

@@

In order to avoid recalculation of the Kirchhoff stress, Algorithm \ref{alg:mf_tensor2} caches its value for each element and quadrature point. In this case, we also avoid the need to evaluate gradients of displacement field with respect to the undeformed configuration.
This algorithm also utilizes the chosen constitutive relationship in that the operation of $J \boldsymbol{\mathcal{C}}$ remains directly expressed using \eqref{eq:simplified_action}.
Therefore, the action of the material fourth-order tangent {\color{red}stiffness tensor} $J \boldsymbol{\mathcal{C}}$ on the second-order symmetric tensor $\overline{\grad^s} \gz x$ is cheap to evaluate.
To that end, we cache two scalars that depend on the Jacobian $J$.

@@

Figures \ref{fig:benchmark_miehe_Emmy} and \ref{fig:benchmark_miehe_IWR} show the results of the matrix-vector multiplication benchmarks as performed on the two clusters.
As expected, the matrix-vector multiplication becomes very expensive for higher order discretization for sparse matrix-based approaches.
The matrix-free implementation is faster than the matrix-based counterpart already for cubic Lagrangian basis in 2D and quadratic in 3D.
Figures \ref{fig:benchmark_miehe_Emmy_vmult2}, \ref{fig:benchmark_miehe_Emmy_vmult3}, \ref{fig:benchmark_miehe_IWR_vmult2} and \ref{fig:benchmark_miehe_IWR_vmult3} clearly show the influence of the different caching strategies.
Two conclusions can be drawn from these results:
The scalar caching implementation is the most time consuming of the three, as at each step we additionally evaluate the gradient of the displacement field in the reference configuration that is required to evaluate the Kirchhoff stress $\gz \tau$.
There is little difference between the approach where the fourth-order material tangent {\color{red}stiffness} tensor is cached (``tensor4'') and the one where we only cache the second-order Kirchhoff tensor and utilize the chosen hyperelastic constitutive equation to efficiently implement the action of the material tangent {\color{red}stiffness tensor} on the second-order symmetric tensor.
This indicates that the time savings we get from the latter approach are insignificant within the complete matrix-free operator implementation.
Consequently, we can apply the matrix-free approach to any finite-strain material model by caching the fourth-order material tangent {\color{red}stiffness tensor} in addition to the Kirchhoff stress and expect this to be faster than the matrix-based implementation.
It is therefore feasible to define highly performant generic operators that are independent of any applied constitutive laws.

>>>>

\subsection{Weak form}
%
% Kinematics
%
The deformation of a body $\mcl B$ from the {\color{red}referential} configuration $\mcl B_0$ to the {\color{red}spatial} configuration $\mcl B_t$ at time $t$
is defined via the deformation map $\gz \varphi_t: \mcl B_0 \rightarrow \mcl B_t$, which places material points of $\mcl B$ into the Euclidean space $\mathbb E^3$.
The current location of a material particle $\gz X$ is given by $\gz x = \gz \varphi_t (\gz X)$.
The displacement field reads $\gz u = \gz x - \gz X$.

@@

%
% strain measures
%
In what follows, we parameterize the material behavior using the right Cauchy-Green tensor $\gz C := \gz F^T \cdot \gz F$.
We will also use the Green-Lagrange strain tensor $\gz E:=\frac{1}{2}\left[\gz C - \gz I\right]$ and the left Cauchy-Green tensor $\gz b := \gz F \cdot \gz F^T$.
Clearly $J = \sqrt{\det \gz C}$ and $\partial (\bullet)/\partial \gz E = 2\, \partial (\bullet)/\partial \gz C$.
%
For conservative systems without body forces, the total potential energy functional $\mcl E$ is introduced as
\begin{align}
\mcl E =
\int_{\mcl B_0} \mcl \psi(\gz F) \, \d V \,
- \int_{\partial \mcl B_0^N} \overline{\gz T} \cdot \gz u \, \d S \, ,
\end{align}
where $\overline{\gz T}$ is the prescribed loading at the Neumann part of the boundary $\partial \mcl B_0^N$ in the {\color{red}referential} configuration, $\mcl \psi$ denotes the strain-energy per unit reference volume and $\gz u$ should satisfy the prescribed Dirichlet boundary conditions $\gz u = \overline{\gz u}$ on $\partial \mcl B_0^D := \partial \mcl B_0 \setminus \partial \mcl B_0^N \neq \emptyset $.

@@

The Newton-Raphson solution approach is adopted for the nonlinear problem considered here.
Given the current trial solution field $\overline{\gz u^h}$, the correction $\Delta \gz u^h$ field is obtained as the solution to the following system of equations (see \eqref{eq:weak_form} and \eqref{eq:tangent}):
\begin{align}
  A_{ij} \Delta u_j &= - F_i  \label{eq:linear_system} \\
  A_{ij} \equiv a(\gz N_i, \gz N_j) &=
  \int_{\mcl B_0}
  \underbrace{
  \overline{\grad^s} \gz N_i : J \boldsymbol{\mathcal{C}} : \overline{\grad^s} \, \gz N_j
  +
  \overline{\grad}\gz N_i :
  \left[
  \overline{\grad} \gz N_j \cdot
  \overline{\gz \tau}
  \right]
  }_{\displaystyle \tilde{a}(\gz N_i, \gz N_j)}
  \d V
  \label{eq:algebraic_tangent}
  \\
  F_i &=
  \int_{\mcl B_0} \overline{\gz \tau} : \grad^{s} \, \gz N_i \, \d V \,
  -
  \int_{\partial \mcl B_0^N} \overline{\gz T} \cdot \gz N_i \, \d S.
\end{align}
Here $\gz A$ is the discrete tangent {\color{red}operator}, $\gz F$ is the discrete gradient
and $\tilde{a}(\gz N_i, \gz N_j)$ is a representation for the integrand in the bilinear form $a(\gz N_i, \gz N_j)$ for the trial solution field $\overline{\gz u^h}$.
Note that $\overline{\grad^s} \gz N_i$ and $\overline{\grad} \gz N_i$ are gradients of shape functions in the {\color{red}spatial} configuration, whereas the integration is done in the {\color{red}referential} configuration.

@@

Given that the matrix-free operators are applied on-the-fly, it is crucial to find a good balance between pre-calculation of certain values (caching) and performance of the algorithm.
Below, we propose three algorithms to implement the tangent {\color{red}operator} of finite-strain elasticity (see \eqref{eq:algebraic_tangent}) using the matrix-free approach.
%
In order to obtain gradients with respect to the {\color{red}spatial} configuration we used the \textit{MappingQEulerian} class from the \texttt{deal.II} \cite{dealII90} library.
In addition to the standard mapping from the reference (isoparametric) element to the element in real space, this class computes the mapping to the {\color{red}spatial} configuration given a displacement field.
%
Due to the specifics of matrix-free operator evaluation implemented in \texttt{deal.II}, the integration is performed over the {\color{red}spatial} configuration. Therefore, we have to divide by the Jacobian $J$ of the deformation map at a given quadrature point additionally to express the integral in the {\color{red}referential} configuration.

@@

Algorithm \ref{alg:mf_scalar} caches a single scalar which involves the logarithm of the Jacobian -- a computationally demanding operation compared to additions and multiplications. As evident from Section \ref{sec:theory}, we need to re-evaluate the Kirchhoff stress at each quadrature point in order to apply the operator. This in turn requires evaluating the deformation gradient $\gz F$ and therefore gradients with respect to the {\color{red}referential} configuration.

@@

In order to avoid recalculation of the Kirchhoff stress, Algorithm \ref{alg:mf_tensor2} caches its value for each element and quadrature point. In this case, we also avoid the need to evaluate gradients of displacement field with respect to the {\color{red}referential} configuration.
This algorithm also utilizes the chosen constitutive relationship in that the operation of $J \boldsymbol{\mathcal{C}}$ remains directly expressed using \eqref{eq:simplified_action}.
Therefore, the action of the material fourth-order tangent {\color{red}stiffness tensor} $J \boldsymbol{\mathcal{C}}$ on the second-order symmetric tensor $\overline{\grad^s} \gz x$ is cheap to evaluate.
To that end, we cache two scalars that depend on the Jacobian $J$.

@@

Figures \ref{fig:benchmark_miehe_Emmy} and \ref{fig:benchmark_miehe_IWR} show the results of the matrix-vector multiplication benchmarks as performed on the two clusters.
As expected, the matrix-vector multiplication becomes very expensive for higher order discretization for sparse matrix-based approaches.
The matrix-free implementation is faster than the matrix-based counterpart already for cubic Lagrangian basis in 2D and quadratic in 3D.
Figures \ref{fig:benchmark_miehe_Emmy_vmult2}, \ref{fig:benchmark_miehe_Emmy_vmult3}, \ref{fig:benchmark_miehe_IWR_vmult2} and \ref{fig:benchmark_miehe_IWR_vmult3} clearly show the influence of the different caching strategies.
Two conclusions can be drawn from these results:
The scalar caching implementation is the most time consuming of the three, as at each step we additionally evaluate the gradient of the displacement field in the {\color{red}referential} configuration that is required to evaluate the Kirchhoff stress $\gz \tau$.
There is little difference between the approach where the fourth-order material tangent {\color{red}stiffness} tensor is cached (``tensor4'') and the one where we only cache the second-order Kirchhoff tensor and utilize the chosen hyperelastic constitutive equation to efficiently implement the action of the material tangent {\color{red}stiffness tensor} on the second-order symmetric tensor.
This indicates that the time savings we get from the latter approach are insignificant within the complete matrix-free operator implementation.
Consequently, we can apply the matrix-free approach to any finite-strain material model by caching the fourth-order material tangent {\color{red}stiffness tensor} in addition to the Kirchhoff stress and expect this to be faster than the matrix-based implementation.
It is therefore feasible to define highly performant generic operators that are independent of any applied constitutive laws.
